# Issue for Decision-Making {.numbered}

## Exploratory Analysis of Multidimensional Data

### Import data set, require packages and a short summary of data

```{r}
#| echo: true
#| message: false
#| warning: false
# working directory
#setwd(dirname(rstudioapi::getSourceEditorContext()$path))

# packages
list_packages = c('readxl', 'dplyr', 'moments', 'tidyr', 'tibble', 'gt', 'ggplot2', 'fmsb')
new.packages = list_packages[!(list_packages %in% installed.packages()[,"Package"])]
if (length(new.packages)) {
  install.packages(new.packages)
}
for (package in list_packages){
  library(package, character.only = T)
}

# Load the dataset
# Define the file path
file_path <- paste0(getwd(),"/delitos_2024_v2.xlsx")

# Load the specific sheet from the Excel file
delitos_data <- read_excel(file_path, sheet = "2024")

dim(delitos_data)
summary(delitos_data)
```

### Skewness

Skewness measures the asymmetry of a data distribution around its mean. It is defined mathematically as:

$[
g_1 = \frac{n}{(n-1)(n-2)} \sum \left( \frac{x_i - \bar{x}}{\sigma} \right)^3
]$

where:

-   $( n )$: Number of observations,
-   $( x_i )$: Individual data points,
-   $( \bar{x} )$: Mean of the data,
-   $( \sigma )$: Standard deviation of the data.

A skewness of $( 0 )$ indicates a perfectly symmetric distribution. Positive skewness $( g_1 > 0 )$ signifies a longer tail on the right side of the distribution, while negative skewness $( g_1 < 0 )$ indicates a longer tail on the left. The code below calculates skewness for all numeric columns in `delitos_data` and presents the results in a formatted table:

```{r}
#| echo: true
#| message: false
#| warning: false

# skewness
delitos_data %>%
  select(where(is.numeric)) %>%  # Select only numeric columns
  summarise(across(everything(), ~ skewness(.x, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Crime Type") %>%
  rename(Skewness = V1) %>%
  gt()
```

### Kurtosis

Skewness measures the asymmetry of a data distribution around its mean. It is defined mathematically as:

$[
g_1 = \frac{n}{(n-1)(n-2)} \sum \left( \frac{x_i - \bar{x}}{\sigma} \right)^3
]$

where:

-   $( n )$: Number of observations,
-   $( x_i )$: Individual data points,
-   $( \bar{x} )$: Mean of the data,
-   $( \sigma )$: Standard deviation of the data.

A skewness of $( 0 )$ indicates a perfectly symmetric distribution. Positive skewness $( g_1 > 0 )$ signifies a longer tail on the right side of the distribution, while negative skewness $( g_1 < 0 )$ indicates a longer tail on the left. The code below calculates skewness for all numeric columns in `delitos_data` and presents the results in a formatted table:

```{r}
#| echo: true
#| message: false
#| warning: false

# kurtosis
delitos_data %>%
  select(where(is.numeric)) %>%  # Select only numeric columns
  summarise(across(everything(), ~ kurtosis(.x, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Crime Type") %>%
  rename(Kurtosis = V1) %>%
  gt()
```

### Variation

Skewness measures the asymmetry of a data distribution around its mean. It is defined mathematically as:

$[
g_1 = \frac{n}{(n-1)(n-2)} \sum \left( \frac{x_i - \bar{x}}{\sigma} \right)^3
]$

where:

-   $( n )$: Number of observations,
-   $( x_i )$: Individual data points,
-   $( \bar{x} )$: Mean of the data,
-   $( \sigma )$: Standard deviation of the data.

A skewness of $( 0 )$ indicates a perfectly symmetric distribution. Positive skewness $( g_1 > 0 )$ signifies a longer tail on the right side of the distribution, while negative skewness $( g_1 < 0 )$ indicates a longer tail on the left. The code below calculates skewness for all numeric columns in `delitos_data` and presents the results in a formatted table:

```{r}
#| echo: true
#| message: false
#| warning: false

# variation
delitos_data %>%
  select(where(is.numeric)) %>%  # Select only numeric columns
  summarise(
    across(
      everything(),
      ~ ifelse(mean(.x, na.rm = TRUE) != 0, 
               sd(.x, na.rm = TRUE) / mean(.x, na.rm = TRUE), 
               NA),  # Compute CV safely
      .names = "{col}"
    )
  ) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Crime Type") %>%
  rename(Variation = V1) %>%
  gt()
```

### Median Absolute Deviation MAD and MAD/median

Skewness measures the asymmetry of a data distribution around its mean. It is defined mathematically as:

$[
g_1 = \frac{n}{(n-1)(n-2)} \sum \left( \frac{x_i - \bar{x}}{\sigma} \right)^3
]$

where:

-   $( n )$: Number of observations,
-   $( x_i )$: Individual data points,
-   $( \bar{x} )$: Mean of the data,
-   $( \sigma )$: Standard deviation of the data.

A skewness of $( 0 )$ indicates a perfectly symmetric distribution. Positive skewness $( g_1 > 0 )$ signifies a longer tail on the right side of the distribution, while negative skewness $( g_1 < 0 )$ indicates a longer tail on the left. The code below calculates skewness for all numeric columns in `delitos_data` and presents the results in a formatted table:

```{r}
#| echo: true
#| message: false
#| warning: false

# MAD
delitos_data %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ mad(.x, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Crime Type") %>%
  rename(MAD = V1) %>%
  gt()

# Standard Deviation (SD)
delitos_data %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ sd(.x, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Crime Type") %>%
  rename(SD = V1) %>%
  gt()

# MAD/Median
delitos_data %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ ifelse(
    median(.x, na.rm = TRUE) != 0,
    mad(.x, na.rm = TRUE) / median(.x, na.rm = TRUE),
    NA
  ))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Crime Type") %>%
  rename(`MAD/Median` = V1) %>%
  gt()

delitos_data %>%
  select(SEMOVIENTES, AUTOMOTORES, BANCOS, PIRATERIA) %>%
  summary()
```

## Graphical analysis

```{r}
#| echo: true
#| message: false
#| warning: false

# Load necessary library
library(ggplot2)
library(tidyr)

# Transform the data to a long format for ggplot
delitos_long <- delitos_data %>%
  select(where(is.numeric)) %>%  # Select only numeric columns
  pivot_longer(cols = everything(), names_to = "Crime Type", values_to = "Value")

# Create faceted histograms
ggplot(delitos_long, aes(x = Value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~ `Crime Type`, scales = "free") +  # Facet by crime type
  theme_minimal() +
  labs(
    title = "Distributions of Crime Data",
    x = "Value",
    y = "Frequency"
  ) +
  theme(
    axis.text.x = element_text(size = 5)  # Reduce the font size of X-axis text
  )
```

Log-transformed crime data distributions

```{r}
#| echo: true
#| message: false
#| warning: false

# Transform the data to long format and apply log transformation
delitos_long_log <- delitos_data %>%
  select(where(is.numeric)) %>%
  mutate(across(everything(), ~ log(.x), .names = "{col}")) %>%  # Log transform (log(x + 1) to avoid log(0))
  pivot_longer(cols = everything(), names_to = "Crime Type", values_to = "Log Value")

# Create faceted histograms for log-transformed values
ggplot(delitos_long_log, aes(x = `Log Value`)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~ `Crime Type`, scales = "free") +  # Facet by crime type
  theme_minimal() +
  labs(
    title = "Log-Transformed Distributions of Crime Data",
    x = "Log Value",
    y = "Frequency"
  ) +
  theme(
    axis.text.x = element_text(size = 3)  # Reduce the font size of X-axis text
  )
```


```{r}
#| echo: true
#| message: false
#| warning: false

# Transform the data to long format and apply log transformation
delitos_long_log <- delitos_data %>%
  select(where(is.numeric)) %>%
  mutate(across(everything(), ~ log(.x), .names = "{col}")) %>%  # Log transform (log(x + 1) to avoid log(0))
  pivot_longer(cols = everything(), names_to = "Crime Type", values_to = "Log Value")

# Create faceted histograms for log-transformed values
ggplot(delitos_long_log, aes(x = `Log Value`)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~ `Crime Type`, scales = "free") +  # Facet by crime type
  theme_minimal() +
  labs(
    title = "Log-Transformed Distributions of Crime Data",
    x = "Log Value",
    y = "Frequency"
  ) +
  theme(
    axis.text.x = element_text(size = 3)  # Reduce the font size of X-axis text
  )
```